---
title: "Machine learning Writeup"
author: "Jan Bezuidenhout"
date: "Tuesday, January 20, 2015"
output: html_document
---

This report illustrates the steps followed to build a predictive model based on the quantified self movement data. 
See <http://groupware.les.inf.puc-rio.br/har> for more details.

Below is a list of the packages use. 
```{r packages, results = 'hide', warning=FALSE, message=FALSE}
library(data.table)
library(rpart)
library(party)
library(partykit)
library(caret)
```

For this report only the training data was used. To reduce the number of features for the model, all features with a variance of less than 10 was excluded.

```{r getdata}
data.Train <- as.data.frame(fread('pml-training.csv'))
data.Train <- subset(data.Train, select=c(magnet_forearm_y,magnet_arm_x,magnet_forearm_z,magnet_forearm_x,magnet_dumbbell_x,
                                          magnet_dumbbell_y,magnet_arm_z,magnet_arm_y,accel_forearm_y,accel_arm_x,accel_forearm_x,
                                          magnet_dumbbell_z,accel_forearm_z,accel_arm_z,accel_arm_y,accel_dumbbell_z,roll_forearm,
                                          yaw_forearm,accel_belt_z,yaw_belt,yaw_dumbbell,accel_dumbbell_y,roll_arm,yaw_arm,
                                          roll_dumbbell,accel_dumbbell_x,magnet_belt_z,magnet_belt_x,roll_belt,pitch_dumbbell,
                                          magnet_belt_y,pitch_arm,accel_belt_x,accel_belt_y,pitch_forearm,pitch_belt,total_accel_arm,
                                          total_accel_dumbbell,total_accel_forearm,total_accel_belt, classe))

data.Train$classe <- as.factor(data.Train$classe)
```

The model was not built with the caret package, but with the rpart package. A Decision tree model was selected because of the ease of interpreting the actual model.

maxdepth was set to 15 although the accuracy of the model started to taper down from 11.

```{r, model}
rpart.model <- rpart(classe ~. ,data = data.Train,  control = rpart.control(maxdepth = 15))
```

Below is an illustration of what the actual model looks like. The plot is not very clear due to the number of nodes selected for this model.

```{r, modelplot}
rpart.plot <- as.party(rpart.model)
plot(rpart.plot)
```

Finally cross validation is performed on one subset of the training data. The estimated accuracy is above 70%. When the actual predictions on the Test set was submitted this model scored 14 out of 20, thus close to the suggested accuracy.

```{r, modelvalidation}
data.test <- data.Train[sample(nrow(data.Train), 300), ]
# summary(data.test$classe)

test.x <- subset(data.test, select=-classe)
test.y <- subset(data.test, select=classe)
test.y <- sapply(test.y, as.factor)
test.y <- as.factor(test.y)
levels(test.y) <- c('A','B','C','D','E')

rpPred <- predict(rpart.model, newdata = data.test, type = 'class')

confusionMatrix(rpPred, test.y)
```

